---
title: "Accessing component databases of VEuPathDB"

author:
- name: "Rohit Satyam, Alberto Maillo, David Gomez-Cabrero, Arnab Pain"
  affiliation: "King Abdullah University of Science & Technology, Saudi Arabia"

abstract: >
  This section covers the capability of `plasmoRUtils` to access various VEuPathDb database and carry out ID conversion tasks, fetch preconfigured data tables and few other tasks. 
clean: false
date: "`r format(Sys.Date(), '%d %B, %Y')`"
package: plasmoRUtils
output: 
  BiocStyle::html_document:
    toc_float: true
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Accessing component databases of VEuPathDB}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BiocStyle} 
  %\VignettePackage{plasmoRUtils}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,cache = TRUE, warning = FALSE,message = FALSE,
  comment = "#>"
)
```

# Introduction

The `r Rpackage("plasmoRUtils")` package streamlines access to VEuPathDB's family of 12 specialized databases such as ToxoDB, PlasmoDB, and PiroplasmaDB [@amos2022veupathdb]. It provides direct data retrieval capabilities through VEuPathDB's RESTful API, enabling seamless integration of biological data into R workflows. The package supports downloading both standard and customized data tables, making it particularly valuable for researchers needing to combine data from multiple sources for downstream analysis.

With release 71 of VEuPathDB, only users with subscription and API key can access the database programmatically. Use `usethis::edit_r_environ(scope = "user")` to add VEuPathDB API using `VEUPATHDB_API_KEY` variable. By default, all the functions look for `VEUPATHDB_API_KEY` variable for API key. Should you store your key using any other variable, supply it using `api_key` argument in VEuPathDB asscoiated functions.

```{r setup}
# Load package and some other useful packages by using
suppressPackageStartupMessages(
  suppressWarnings({
    library(plasmoRUtils)
    library(dplyr)
    library(plyr)}))
```

# Gene ID Conversion

A common challenge in bioinformatics involves mapping between different identifier systems across databases. For apicomplexan research, this might include converting between UniProt IDs, legacy gene identifiers, and current Ensembl gene IDs. The `toGeneid()` function addresses this need by retrieving up-to-date annotations from VEuPathDB databases, supporting bidirectional conversion between various ID types through flexible parameter specification.

## Retrieving Gene Annotations and Alternative IDs

The `toGeneid()` function enables annotation retrieval when provided with Ensembl gene IDs. By default, it returns essential information including gene names, symbols, outdated gene IDs, and protein Uniprot IDs. The function's versatility extends to supporting custom field requests through its `customFields` parameter, with available options documented in the `getTable()` help section.

```{r}
## Get annotations for list of geneIDs for PF3D7
toGeneid(c("PF3D7_0420300", "PF3D7_0621000"), from="ensembl")
## Get annotations for list of geneIDs for organisms other than PF3D7
toGeneid(inputid = c("TGME49_304740","TGME49_208030"),from="ensembl",org="Toxoplasma gondii ME49", db="toxodb")

## Convert uniprot IDs back to gene IDs. It will also provide Product description and Gene Symbol

toGeneid(inputid = c("Q8I1N6","C6KT48"),from="uniprot",to="ensembl" )


## Using customFields to get only columns of interest
toGeneid(inputid = c("TGME49_304740","TGME49_208030"),
         from="ensembl",org="Toxoplasma gondii ME49",
         db="toxodb",
         customFields=c("primary_key","predicted_go_component","annotated_go_function"))

```

> **Note:** Successful ID conversion requires precise organism nomenclature matching VEuPathDB's conventions. For example, *`Toxoplasma gondii ME49`* must include proper spacing and special characters. Invalid query would be *`Toxoplasma gondiiME49`* or short forms `TgME49`.

This functionality proves particularly valuable for enhancing differential expression analysis results with comprehensive annotations, enabling complete workflow automation without leaving command-line interfaces on HPC systems.

With `toGeneid()` function, one-to-many relationships are retained in long format by default, allowing all possible mappings to remain visible. The decision about how to handle such cases is therefore left to the user, rather than being imposed by the function. 

For instance, the two protein IDs `Q8I0P6` and `A0A2P1JI60` correspond to `elongation factor 1-alpha`. Although the genes `PF3D7_1357100` and `PF3D7_1357000`  encode the same protein (thereby same uniprot ID `Q8I0P6`), they are located on opposite strands: `PF3D7_1357100` is on the forward strand, whereas `PF3D7_1357000` is on the reverse strand. They also differ in transcript length. In this example if the task is Uniprot to Gene ID mapping, users can group the results returned by `toGeneid()` function by Uniprot IDs and collapse rows thereby concatinating two gene IDs separated by semicolon or comma. An alternative is to randomly choose one lexicographically. Because there is no universally appropriate rule for resolving this type of ambiguity, we avoid making this choice for the end user. 

```{r}
toGeneid(c("Q8I0P6" ,"A0A2P1JI60"), from = "uniprot",to = "ensembl")
```


## Accessing Preconfigured Data Tables from VEuPathDB's component sites

Since VEuPathDB API documentation specifically encourages to use specific organism database as quoted below, we developed `getTable()` function to fetch fields of interests from each database separately.

> There are 12 component sites and one portal: VEuPathDB.org. The component sites are: AmoebaDB, CryptoDB, FungiDB, GiardiaDB, HostDB, MicrosporidiaDB, PiroplasmaDB, PlasmoDB, ToxoDB, TrichDB, TriTrypDB and VectorBase. For most record types (all but dataset and organism), when running a search, the portal reaches out to component sites to get the search results. That means it will be faster to use a component site directly when you can.

Some frequently required fields have been provided in the help section of `getTable()`.

```{r}
## To fetch table for all the genes present in an organism

getTable(org="Plasmodium falciparum 3D7", db="plasmodb",
         api_key = Sys.getenv("VEUPATHDB_API_KEY")) %>% head()

## User can also provide custom fields. For example we wish to download the Transcript abundance in Transcripts per Million (TPM) [Dataset: Transcriptome in severe vs uncomplicated malaria]

getTable(org="Plasmodium falciparum 3D7", db="plasmodb", customFields = c("primary_key","pan_24905_ns_2917")) %>% head()

## The files to be supplied to customFields arguments can be found using
attrs <- getTableAttributes(
   db = "plasmodb",
  record_type = "transcript", api_key=Sys.getenv("VEUPATHDB_API_KEY")
)

head(attrs)

```

For more information on fields that can be supplied to `getTable()`, use the following steps:

1.  Go to database of your interest (Say "PlasmoDB")

2.  Click on `Annotation, curation and identifiers` tab on your left and select `List of IDs`

3.  Scroll down and click on `Build a Web Services URL from this Search >>` hyperlink.

4.  Under the section `Choose Columns:` choose fields of your interest. Most of these fields are included in help section of `getTable()`. However, fields specific to a particular database such as dataset related fields (starts with "pan\_") are excluded.

5.  Once you select fields, they are updated in **POST** section of the webpage query builder.

The constituent databases of VEuPathDB also provide some preconfigured tables which can not be fetched via `getTable()` function. To enable users to fetch such tables, we wrote another function called `getPreconfiguredTable()` the usage of which has been shown below.

```{r}
## To get list of preconfigured tables

preconf <- getPreconfiguredTable(listtables = TRUE)

head(preconf)

## Fetch pathway table for all the genes from MPMP database

getPreconfiguredTable(org = "Plasmodium falciparum 3D7",db = "plasmodb",customField = "MetabolicPathwaysMPMP") %>% head()

```

Please note that the MPMP pathway version provided by PlasmoDB is [outdated](https://plasmodb.org/plasmo/app/record/dataset/DS_1e177b728b) (`03-2019`). Some pathways have been revised or removed in its entirety. If you wish to access the latest MPMP version, you can use `data("mpmp.28Aug2024")` for you analysis which was scraped by us. If you wish to use this geneset for MPMP pathway enrichment analysis using `r CRANpkg("pathfindR")`, you can do so by using `data("pathfindrMPMP")`.

> Note: We urge the users to cite the original articles of the related datasets alongside plasmoRUtils.

## Fetching genome metadata and strain names

In above examples, we saw the importance of passing exact name to the `org` argument for `toGeneid()` to function properly. A helper function is provided to achieve this called `listVeupathdb()`. By default, 11 columns are returned including organism name, the respective database present in "VEuPathDB Project" column and some more additional information. However, you can limit the search to columns of your interests as shown below. As stated by VEuPathDB:

> The best use of the VEuPathDB portal is to get a table with all organisms in our sites, and for each organism: the component site, and the urls to access their fasta and gff files.

```{r}
listVeupathdb() %>% head()
listVeupathdb(customFields=c("species", "project_id")) %>% head()
```

## Mapping PDB IDs to Gene IDs

There is currently no facility in VEuPathDB to convert the PDB IDs to respective gene IDs. If PDB ID corresponds to a multimer complex and if you have multiple such PDB ids, it becomes arduous to map them to Gene IDs manually. To provide solution to this issue, we can first convert the PDB chains to Uniprot IDs using our own `pdb2uniprot()` function and then can use `toGeneid()` function to obtain gene IDs.

```{r}
pdbids <- c("7D2W","4U5A","6E10")
df <- lapply(pdbids, pdb2uniprot) %>% plyr::ldply()

geneids <- toGeneid(inputid = unique(df$attribute),from = "uniprot",to = "ensembl")

## Combining the geneIDs with df
S4Vectors::merge(df,geneids,all=TRUE, by.x="attribute", by.y="UniProt ID(s)") %>% head()

```

# Session {.unnumbered}

```{r}
sessionInfo()
```

# References {.unnumbered}
