---
title: "Retrieving Orthologs"

author:
- name: "Rohit Satyam, Alberto Maillo, David Gomez-Cabrero, Arnab Pain"
  affiliation: "King Abdullah University of Science & Technology, Saudi Arabia"
abstract: >
  This is a quick and dirty tutorial on how to fetch orthologs from OrthoMCL and InparanoiDB databases.
date: "`r format(Sys.Date(), '%d %B, %Y')`"
package: plasmoRUtils
output: 
  BiocStyle::html_document:
    toc_float: true
    toc: true
    number_sections: true
vignette: |
  %\VignetteIndexEntry{Retrieving Orthologs}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BiocStyle}
  %\VignettePackage{plasmoRUtils}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,cache = TRUE, warning = FALSE,message = FALSE,
  comment = "#>"
)
```

## Introduction

Users might require Ortholog mapping to:

-   to assess phyletic profile of a set of genes and see if they have orthologs in other related organism or not.

-   integrate datasets obtained from two different species. Often such datasets are scRNASeq (eg: [@Tebben2022])

-   borrow annotations such as localization for hypothesis generation.

-   perform co-expression analysis between two different species.

Whatever the use case maybe, we will see how we can exploit `getpairedOrthologs()` to fetch orthologs of multiple organisms of interest in batch manner and integrate them.

With release 71 of VEuPathDB, only users with subscription and API key can access the database programmatically. Use `usethis::edit_r_environ(scope = "user")` to add VEuPathDB API using `VEUPATHDB_API_KEY` variable. By default, all the functions look for `VEUPATHDB_API_KEY` variable for API key. Should you store your key using any other variable, supply it using `api_key` argument in VEuPathDB asscoiated functions.

```{r setup}
# Load package and some other useful packages by using
suppressPackageStartupMessages(
  suppressWarnings({
    library(plasmoRUtils)
    library(dplyr)
    library(plyr)}))
```

Here, we will try to get all *Plasmodium falciparum* 3D7 orthologs present in *Toxoplasma gondii* ME49, *Plasmodium berghei* ANKA and *Plasmodium vivax* Sal1. 

```{r}
#?getpairedOrthologs
ids <- listOrthomcl()
numids <- c(ids %>% dplyr::filter(Organism=="Plasmodium vivax Sal-1") %>% dplyr::pull(ID),
            ids %>% dplyr::filter(Organism=="Toxoplasma gondii ME49") %>% dplyr::pull(ID),
            ids %>% dplyr::filter(Organism=="Plasmodium berghei ANKA") %>% dplyr::pull(ID))
res <- lapply(numids, function(x){
  getpairedOrthologs(from = ids %>% dplyr::filter(Organism=="Plasmodium falciparum 3D7") %>% dplyr::pull(ID),
                     to=x,
                     db="orthomcl",transform = FALSE)
}) %>% setNames(c("Pvivax","Tgme49","Pb"))

merged_df <- Reduce(function(x, y) merge(x, y, by = "Accession", all = TRUE), res)
merged_df %>% head()

## Tidying up the dataframe by removing organisms prefixes
strip_prefix <- function(x) sub("^[^|]+\\|", "", x)     # per-token
clean_list   <- function(s) {
    toks <- trimws(strsplit(s, ",", fixed = TRUE)[[1]])
    paste(vapply(toks, strip_prefix, character(1)), collapse = ",")
}

merged_df[] <- lapply(merged_df, function(col) vapply(col, clean_list, character(1)))

## Changing column names
colnames(merged_df) <- c("P falciparum 3D7","P vivax Sal1","T gondii ME49","Plasmodium berghei ANKA")

merged_df %>% head()
```

Since OrthoMCL and other VEuPathDB database are not updated simultaneously, chances are that you might be using old OrthoMCL ID. If that's the case above function will not be much helpful. The workaround is to fetch the old OrthoMCL IDs for each organisms and their respective databases and use it as an anchor to combine and collapse gene IDs from all the organisms. The code snippet below demonstrate the same.

```{r}
## List all organisms
list <- listVeupathdb(customFields = c("primary_key","project_id","species",'species_ncbi_tax_id'))

## Subset organisms I am interested in 
dbs <- list[grep("3D7|ME49$|Sal-1", list$Organism),]
dbs
df <- lapply(1:nrow(dbs), function(x){
  plasmoRUtils::getTable(org=dbs[x,]$Organism, db=tolower(dbs[x,]$project_id),customFields = c("primary_key" ,"gene_orthomcl_name"))
})

## Retain protein coding genes

df2 <- lapply(df, function(x){
  x[!(stringr::str_detect(pattern = "N/A",string = x$`Ortholog Group`)),]
})

## Combine all the tables
merged_df2 <- Reduce(function(x, y) merge(x, y, by = "Ortholog Group", all = TRUE), df2)

merged_df2 %>% head(n = 10)
## Combining all the gene IDs in each column so that we have 1 row per orthogroup
merged_df2 <- merged_df2 %>% 
  dplyr::group_by( `Ortholog Group`) %>% 
  dplyr::summarise(dplyr::across(dplyr::everything(), 
                                 ~paste(unique(.x), collapse = ",")), .groups = "drop")

## Changing column names
colnames(merged_df2) <- c("Orthogroup ID","P vivax Sal1","P falciparum 3D7","T gondii ME49")
merged_df2 %>% head(n = 10)
```

## Fetching orthologs from InParanoiDB

InParanoiDB uses NCBI taxon IDs as unique identifiers. For *P. falciparum* (36329), *T. gondii* (5811), *P. vivax* (126793) and *P. berghei* (5823), we will fetch pairwise orthologs by keeping *P. falciparum* as constant species. We can recycle the code above as follows.

```{r}
#?getpairedOrthologs
# listipdb()

res <- lapply(c(126793,5811,5823), function(x){
  getpairedOrthologs(from = 36329,
                     to=x,
                     db="ipdb",transform = TRUE)
}) %>% setNames(c("Pvivax","Tgme49","Pb"))

res[[1]] %>% head()

```

Combining the results across InParanoiDB, is not as straightforward as for OrthoMCL because of lack of unique Orthogroup ID. However, user can use Query Uniprot ID as anchor, combine the results, convert Uniprot IDs to gene IDs and then collapse rows since two or more Uniprot IDs might correspond to same gene IDs.

## Fetching Preconfigured tables from OrthoMCL-DB and map old OG IDs to new IDs

OrthoMCL gets frequently updated and with every update come new Ortholog IDs. When using published dataset, its highly likely that users would want to map the old orthogroups to new orthogroups or assess which proteins changed the orthogroups and got clubbed, which orthogroup got split etc. `getPreconfiguredTableOrthomcl` can help you do this by accessing preconfigured table used by OrthoMCL to do that.
```{r}
## Fetching all OG IDs that maps to IDs of interest

## New to old IDs. Old Ids are present in second and third column
ids = c("OG7_0008348", "OG7_0003896")
getPreconfiguredTableOrthomcl(ids,customField = "previousGroups")

## Old Id. New IDs are present in first column
getPreconfiguredTableOrthomcl("OG3_10277")
```

By changing `customField` arguments, you can also access other preconfigured tables from the database. Eg. if you want keyword frequencies and their associated orthogroups, this can be achieved using

```{r}
getPreconfiguredTableOrthomcl(customField = "DomainFrequency")
```

# Session Info {.unnumbered}

```{r}
utils::sessionInfo()
```

# References {.unnumbered}